---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file, then run devtools::build_readme(). -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%"
)
```

# cevcmm

<!-- badges: start -->
[![R-CMD-check](https://github.com/lidajalili/cevcmm/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/lidajalili/cevcmm/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/lidajalili/cevcmm/branch/main/graph/badge.svg)](https://app.codecov.io/gh/lidajalili/cevcmm?branch=main)
[![CRAN status](https://www.r-pkg.org/badges/version/cevcmm)](https://CRAN.R-project.org/package=cevcmm)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

**cevcmm** fits **Varying Coefficient Mixed-Effects Models (VCMMs)** at scale, including settings where data is split across multiple machines or doesn't fit in memory. It implements the sufficient-statistics (SS) and one-step communication-efficient surrogate-likelihood (CSL) estimators of Lin and Jalili (2026), with diagonal, Kronecker, and separable random-effect covariance structures.

The model is

$$
y_i \;=\; \beta_0(t_i) \;+\; \sum_{k=1}^{K} x_{ik}\,\beta_k(t_i) \;+\; z_i^\top \alpha \;+\; \varepsilon_i,
$$

where each $\beta_k(t)$ is a smooth function of $t$ (cubic B-splines with a second-order-difference penalty), $\alpha \sim N(0, \Sigma_\alpha)$, and $\varepsilon_i \sim N(0, \sigma_\varepsilon^2)$.

## Installation

You can install the released version of cevcmm from CRAN with:

```r
install.packages("cevcmm")
```

Or the development version from GitHub:

```r
# install.packages("remotes")
remotes::install_github("lidajalili/cevcmm")
```

## Quick example

A 500-observation fit with one varying coefficient and three diagonal random effects:

```{r example}
library(cevcmm)

set.seed(1)
N <- 500L
q <- 3L

t <- runif(N)
x <- runif(N)
Z <- matrix(rnorm(N * q), N, q)
y <- 2 + sin(2 * pi * t) * x +
     as.vector(Z %*% rnorm(q, sd = 0.4)) +
     rnorm(N, sd = 0.5)

fit <- vcmm(y, X = x, Z = Z, t = t,
            control = vcmm_control(sigma_eps       = 0.5,
                                   sigma_alpha     = 0.4,
                                   update_variance = TRUE))
fit
```

All standard S3 methods work on the result: `coef()`, `fixef()`, `ranef()`, `vcov()`, `nobs()`, `logLik()`, `AIC()`, `BIC()`, `predict()`, `summary()`, and `plot()`.

## Three vignettes

Three vignettes ship with the package, covering the main use cases end-to-end:

- **`vignette("getting-started", package = "cevcmm")`** — single-machine fit on a simple diagonal-random-effects example. Walks through every public S3 method.
- **`vignette("distributed-fitting", package = "cevcmm")`** — split data across $K$ nodes, compute per-node summaries with `node_summary()`, and recover a fit bit-equivalent to pooled `vcmm()` via `fit_from_summaries()`. Also covers the streaming-accumulator pattern for memory-constrained settings.
- **`vignette("od-migration", package = "cevcmm")`** — Kronecker covariance for origin-destination flow data, using a bundled simulated migration dataset.

## Citation

If you use cevcmm in published work, please cite:

> Lin, L.-H. and Jalili, L. (2026). *Scalable and Communication-Efficient Varying Coefficient Mixed-Effects Models: Methodology, Theory, and Applications.* Journal of the American Statistical Association (under review).

BibTeX:

```bibtex
@article{lin2026cevcmm,
  title   = {Scalable and Communication-Efficient Varying Coefficient
             Mixed-Effects Models: Methodology, Theory, and Applications},
  author  = {Lin, Li-Hsiang and Jalili, Lida},
  journal = {Journal of the American Statistical Association},
  year    = {2026},
  note    = {Under review}
}
```

## License

GPL-3. See the [`LICENSE`](LICENSE) file for the full text.
