---
title: "Distributed fitting with cevcmm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distributed fitting with cevcmm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
set.seed(23)
```

## When you need this

Three scenarios call for the distributed API:

1. **Data lives on $K$ different machines** — privacy or governance constraints
   prevent pooling, but each site can publish a small summary.
2. **Data is on one machine but doesn't fit in memory** — process chunk by
   chunk and accumulate.
3. **Streaming data** — new batches arrive over time and refitting from
   scratch is expensive.

All three are solved with the same primitive: each chunk produces a
`vcmm_ss` summary; summaries are additive; the fit is computed from the
aggregate.

## The math guarantee

Theorem 1 of Lin and Jalili (2026): partition the data into $K$ chunks,
compute a sufficient-statistics summary on each chunk, sum the summaries,
and fit. The result is **bit-equivalent** (up to BLAS summation noise) to
fitting on the pooled data. The summaries lose no information — they
*are* the sufficient statistics for the VCMM likelihood.

The summary on chunk $k$ stores
$$
\bigl(a_k,\; b_k,\; C_k,\; (Z^\top Z)_k,\; (Z^\top y)_k,\; (X^\top Z)_k,\; n_k\bigr)
$$
where $a = y^\top y$, $b = X^\top y$, $C = X^\top X$. Adding summaries
across chunks is element-wise addition; the aggregate is the same as the
single summary you would have computed on the pooled data.

## Setup

```{r setup}
library(cevcmm)
```

## Simulate

A modest example: $N = 1200$ observations, $K = 1$ varying covariate,
$q = 4$ independent random effects. Then we'll split it into 3 nodes.

```{r simulate}
set.seed(23)
N <- 1200L
q <- 4L

t <- runif(N)
x <- runif(N)
Z <- matrix(rnorm(N * q), N, q)

beta_1_fun <- function(u) sin(2 * pi * u)
alpha_true <- rnorm(q, sd = 0.5)
y <- 2 + beta_1_fun(t) * x + as.vector(Z %*% alpha_true) +
     rnorm(N, sd = 0.5)
```

## Pattern 1 — list of node summaries

This is the "data on $K$ machines" case. Each node:

1. computes the same spline design matrix on its own slice of $t$ and $X$,
2. calls `node_summary()` on its slice,
3. ships the resulting `vcmm_ss` object (small — a few KB regardless of $n_k$)
   to the central node.

The central node sums the summaries and calls `fit_from_summaries()`.

```{r split}
# Assign each observation to one of 3 nodes
node_id <- sample.int(3L, N, replace = TRUE)
splits  <- split(seq_len(N), node_id)
lengths(splits)
```

Build the spline design once. (In a real deployment, each node would
construct its piece of the design locally using the same `n_basis`,
`degree`, and knot placement — the package handles this automatically as
long as the inputs agree.)

```{r design}
design <- build_vcmm_design(X = x, t = t)
X_d    <- design$X_design
dim(X_d)   # N x p, where p = 12 = intercept + 11 spline coefs
```

Each node produces a summary:

```{r node-summaries}
summaries <- lapply(splits, function(idx) {
  node_summary(y[idx], X_d[idx, , drop = FALSE], Z[idx, , drop = FALSE])
})
summaries[[1]]
```

The size of each summary is independent of $n_k$:

```{r size}
object.size(summaries[[1]])
object.size(y[splits[[1]]]) + object.size(X_d[splits[[1]], ])
```

Now fit from the summaries — pass the list directly:

```{r fit-dist}
ctrl <- vcmm_control(sigma_eps       = 0.5,
                     sigma_alpha     = 0.5,
                     update_variance = TRUE)

fit_dist <- fit_from_summaries(summaries,
                               penalty = design$penalty,
                               control = ctrl,
                               method  = "ss",
                               re_cov  = "diag")
fit_dist
```

## Verify bit-equivalence against pooled fitting

```{r fit-pooled}
fit_pooled <- vcmm(y, X = x, Z = Z, t = t,
                   method = "ss", re_cov = "diag",
                   control = ctrl)

# beta estimates
max(abs(fit_pooled$beta - fit_dist$beta))

# alpha estimates
max(abs(fit_pooled$alpha - fit_dist$alpha))

# variance components
all.equal(fit_pooled$sigma_eps,   fit_dist$sigma_eps)
all.equal(fit_pooled$sigma_alpha, fit_dist$sigma_alpha)
```

The maximum coefficient difference is at the BLAS summation-noise level
(typically $\sim 10^{-12}$). The distributed fit *is* the pooled fit; only
the path that produced it differs.

## Pattern 2 — streaming accumulator

When you have many small chunks and don't want to keep them all in
memory, use the accumulator instead of a list. The accumulator stores a
running sum of the sufficient statistics; chunks can be incorporated and
discarded as they arrive.

```{r accumulator}
p <- ncol(X_d)
acc <- init_accumulator(p = p, q = q)

# Pretend chunks arrive one at a time
for (k in seq_along(splits)) {
  idx <- splits[[k]]
  ss_k <- compute_sufficient_stats(y[idx],
                                   X_d[idx, , drop = FALSE],
                                   Z[idx, , drop = FALSE])
  acc <- accumulate_stats(acc, ss_k)
}
acc
```

Fit from the accumulator — same call signature:

```{r fit-acc}
fit_acc <- fit_from_summaries(acc, penalty = design$penalty,
                              control = ctrl, method = "ss", re_cov = "diag")

# Same answer as Pattern 1
max(abs(fit_acc$beta - fit_dist$beta))
```

## Identifiability for OD designs

When `rowSums(Z)` is constant — for example, a row of $Z$ has exactly one
"origin" indicator and one "destination" indicator, so each row sums to 2
— the model has a one-dimensional identifiability redundancy: shifting
every $\alpha_j$ by a constant $c$ and subtracting $\sum_j \alpha_j$
adjustments from $\beta_0$ leaves the likelihood unchanged.

`vcmm()` detects this automatically and applies a centering shift so
$\sum_j \hat\alpha_j = 0$. `fit_from_summaries()` cannot detect it from
the summaries alone (the row-sum information is lost when you collapse to
$X^\top Z$), so you have to pass it in via `rowsum_constant`:

```{r od-shift, eval = FALSE}
fit_dist_od <- fit_from_summaries(
  summaries, penalty = design$penalty, control = ctrl,
  method = "ss", re_cov = "kronecker",
  n_groups = G,
  Sigma_spatial_init = Sigma_spatial,
  rowsum_constant    = 2     # <-- this matches what vcmm() applies
)
```

Without `rowsum_constant`, the distributed fit's $\hat\beta_0$ and
$\hat\alpha$ differ from the pooled fit's by exactly the centering
shift. The fits are the same up to the redundancy; the choice of
representative within the equivalence class is what differs.

See `vignette("od-migration", package = "cevcmm")` for a full OD example.

## Where to go next

- **OD migration with Kronecker covariance**:
  `vignette("od-migration", package = "cevcmm")`.
- **Basic single-machine usage**:
  `vignette("getting-started", package = "cevcmm")`.

## Reference

Lin, L.-H. and Jalili, L. (2026). *Scalable and Communication-Efficient
Varying Coefficient Mixed-Effects Models: Methodology, Theory, and
Applications.* JASA (under review).
